Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-06-06 sports_analytics.Rmd
added section on probability of scoring one run

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: nnguyen22@luc.edu
  - name: Katherine Evans
    affiliation: Monumental Basketball
    location: Washington, DC 20003
    email: kevans@monumentalsports.com
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data. Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights. This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   # use apa for WIRE
output: rticles::arxiv_article
---

@@

```{r pkg}
knitr::write_bib(
  c(
    "Lahman", "pitchRx", "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2"
  ),
  file = "pkgs.bib"
)
```

@@

- [CRAN Task View SportsAnalytics](https://github.com/beanumber/ctv-sportsanalytics)

# The expected value of a possession

@@

See @Albert2001curve for a fuller discussion of the use of the expected run matrix in baseball and @marchi2018analyzing for examples of how to estimate the expected run matrix using Retrosheet data.

@@

>>>>

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: nnguyen22@luc.edu
  - name: Katherine Evans
    affiliation: Monumental Basketball
    location: Washington, DC 20003
    email: kevans@monumentalsports.com
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data. Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights. This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   # use apa for WIRE
output: rticles::arxiv_article
header-includes:
  - \usepackage{amsmath}
---

@@

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", "pitchRx", "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2"
  ),
  file = "pkgs.bib"
)
```

@@

See @cochran2017oxford and @Albert2017handbook for overview readers.

See @baumer2022ctv for R packages useful for sports analytics.

# The expected value of a possession {#sec:ev}

@@

See @Albert2001curve for a fuller discussion of the use of the expected run matrix in baseball and @marchi2018analyzing for examples of how to estimate the expected run matrix using Retrosheet data and the R statistical computing language [@R-base].

@@

## Optimal strategies that don't maximize expected points

In Section \ref{sec:ev}, we defined the expected value of a possession based on the state $s$ of the game in terms of the expected number of points (runs) $X$ that would be scored in the remainder of some period of time. 
We then showed how this value could be used to analyze the relative effectiveness of certain strategies, with the simple idea that strategies that yield higher expected values are preferable. 
Generally, the goal of any sport is to score more points than the other team, which most often means trying to score as many points as possible, leading to a general strategy of maximizing expected points. 
However, there are situations in which maximizing the number of expected points is *not* the desired strategy.

For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs. 
Mathematically, we distinguish between 
$$
  \max{\Pr[X > 0 | s]} \, \text{ and } \, \max{\mathbb{E}[X | s]} \,.
$$
Maximizing the probability of scoring is optimal in any sudden-death situation, which has (but currently does not) include overtime in American football [@martin2018markov].

The situation gets even more interesting when teams modify both their offensive and defensive strategies simultaneously. 
For example, in hockey teams will often pull their goalie when trailing in the final period. 
This strategy severely weakens their defense, but strengthens their offense. 
The hope is to score a quick goal to get back in the game, but the risk is falling further behind. 
@beaudoin2010strategies show that NHL coaches do not always employ the optimal strategies. 
@skinner2011scoring develops a general framework for these desperation strategies, which include the onside kick in American football, pulling the infield and/or outfield in in baseball, and of course, the fabled Hack-a-Shaq strategy in basketball.
